---
output: html_document
editor_options:
  chunk_output_type: console
---
# Exploratory Data Analysis {#exploratory-data-analysis .r4ds-section}

## Introduction {#introduction-3 .r4ds-section}

```{r message=FALSE}
library("tidyverse")
library("viridis")

library("nycflights13")

# This is a cool library
library("lvplot")
```


### Exercise 7.3.1 


Explore the distribution of each of the `x`, `y`, and `z` variables in `diamonds`. What do you learn?

Get the summary of x, y, z
```{r}
summary(select(diamonds, x, y, z))
```

Plot the distributions of each
```{r}
ggplot(diamonds) +
  geom_histogram(aes(x = x), binwidth = .2)
```

```{r}
ggplot(diamonds) +
  geom_histogram(aes(x = y), binwidth = .2)
```

```{r}
ggplot(diamonds) +
  geom_histogram(aes(x = z), binwidth = .2)
```
Summary: 

1. x and y contains values are larger with IQR between 4.7 and 6.5
2. while z has IQR between 2.9 and 4.0

Observations:
1. There some outliers in the x, y, z variables which cause skewing to the right
2. There are spikes in the plots, as opposed to a smooth distribution, which suggests multi-modal data.
3. The outliers are either 0 or have extreme values (data entry errors?).

Investigate the outliers:
```{r}
filter(diamonds, x == 0 | y == 0 | z == 0)
```

1. The values with `z` as 0 would be flat, which makes no sense for how diamonds are cut.

```{r}
diamonds %>%
  arrange(desc(x)) %>%
  head()
```

1. Other than the 0 in the low end...the high end of values appears normal for `x`

```{r}
diamonds %>%
  arrange(desc(y)) %>%
  head()
```

1. In the high end of values, `y` has a couple of large outlier values of `58.9` and `31.8`

```{r}
diamonds %>%
  arrange(desc(z)) %>%
  head()
```

1. In the high end of values, `z` has  a single outlier values of `31.8`

Let's plot some comparisons to the x, y, and z values to see how the outliers affect distributions
```{r}
ggplot(diamonds, aes(x = x, y = y)) +
  geom_point()
```

```{r}
ggplot(diamonds, aes(x = x, y = z)) +
  geom_point()
```

```{r}
ggplot(diamonds, aes(x = y, y = z)) +
  geom_point()
```

Let's remove the outliers to see if the relationships make more sense
```{r}
diamonds %>%
filter(x > 0, x <= 10) %>%
  ggplot() +
  geom_histogram(mapping = aes(x = x), binwidth = .01) +
  scale_x_continuous(breaks=1:10)
```

```{r}
diamonds %>%
filter(y > 0, y <= 10) %>%
  ggplot() +
  geom_histogram(mapping = aes(x = y), binwidth = .01) +
  scale_x_continuous(breaks=1:10)
```

```{r}
diamonds %>%
filter(z > 0, z <= 10) %>%
  ggplot() +
  geom_histogram(mapping = aes(x = z), binwidth = .01) +
  scale_x_continuous(breaks=1:10)
```

Think about a diamond and how you might decide which dimension is the length, width, and depth.
```{r}
?diamonds
```

Docs say that 
1. x = length, y = width, z = depth
2. However there's also a `depth` percentage column where it's expressed as:
z / mean(x,y)

```{r}
????
```

### Exercise 7.3.2 


Explore the distribution of price. Do you discover anything unusual or surprising? (Hint: Carefully think about the `binwidth` and make sure you try a wide range of values.)

```{r}
summary(select(diamonds, price))
```
More diamonds are being sold around `$2400`. 

```{r}
diamonds %>%
  filter(price < 2400) %>%
  ggplot() +
  geom_histogram(aes(x = price),binwidth = 10)
```

```{r}
diamonds %>%
  filter(price < 2400) %>%
  ggplot() +
  geom_histogram(aes(x = price),binwidth = 25)
```

```{r}
diamonds %>%
  filter(price < 2400) %>%
  ggplot() +
  geom_histogram(aes(x = price),binwidth = 100)
```
There's also a large drop in qty in sales of diamonds in the data around `$1500`


??? Distribution of last digit?


### Exercise 7.3.3 

How many diamonds are 0.99 carat?
How many are 1 carat?
What do you think is the cause of the difference?

```{r}
diamonds %>%
  filter(carat == .99) %>%
  count(carat)
```
`23` diamonds are .99 carat.

```{r}
diamonds %>%
  filter(carat == 1.0) %>%
  count(carat)
```
`1558` diamonds are 1 carat.

I'm not a diamond cutter but perhaps the 1 carat diamond is a more disarable cut, or perhaps the .99 where diamonds that has a problem during the cut that required further correction that reduced the weight (carat size) of the diamond?

```{r}
diamonds %>%
  filter(carat >= 0.9, carat <= 1.1) %>%
  count(carat) %>%
  print(n = Inf)
```

## Covariation 

#### Exercise 7.5.1.1 

Use what you've learned to improve the visualization of the departure times of cancelled vs. non-cancelled flights.

```{r}
flights = nycflights13::flights

flights %>%
  mutate(
    cancelled = is.na(dep_time), # x = T/F was cancelled
    hour_dep = sched_dep_time %/% 100,
    min_dep = sched_dep_time %% 100,
    sched_dep_time = hour_dep + min_dep / 60 # y = convert time to real number
  ) %>%
  ggplot() +
    geom_boxplot(aes(x=cancelled,y=sched_dep_time))
```


#### Exercise 7.5.1.2


What are the general relationships of each variable with the price of the diamonds?

Carat vs. Price (Continuous variables)
```{r}
diamonds %>%
ggplot() +
  geom_point(aes(x = carat, y=price))
```

```{r}
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_boxplot(mapping = aes(group = cut_width(carat, 0.1)))
```

Color vs Clarity (Boxploting discrete variables)
Use fct_rev to order from worst color to best. (J down to D)
Plost will show weak negative relationship between price and color
```{r}
diamonds %>%
  mutate(color = fct_rev(color)) %>%
  ggplot(aes(x = color, y = price)) +
  geom_boxplot()
```

Carity vs Price (Boxplot discrete vs continuous variables)
Contiuous variable already ordered from worst to best clarity ()
```{r}
ggplot(data = diamonds) +
  geom_boxplot(mapping = aes(x = clarity, y = price))

# not sure how IF sorts after all other values beginning with S anvd V
# I would think I1, IF, then .... for ascending order
diamonds %>%
  group_by(clarity) %>%
  summarize(n = n()) %>%
  arrange(clarity)
```


#### Exercise 7.5.1.4 

One problem with box plots is that they were developed in an era of much smaller datasets and tend to display a prohibitively large number of "outlying values".
One approach to remedy this problem is the letter value plot.
Install the lvplot package, and try using `geom_lv()` to display the distribution of price vs cut.

An example of the plot is shown below (`geom_lv`). Write answers (not code) to answer these questions:
What do you learn?
????

How do you interpret the plots?
????


```{r}
ggplot(diamonds, aes(x = cut, y = price)) +
  geom_lv()

ggplot(diamonds, aes(x = cut, y = price)) +
  geom_boxplot()
```



#### Exercise 7.5.2.2 

Use `geom_tile()` together with dplyr to explore how average flight delays vary by destination and month of year.
What makes the plot difficult to read?
How could you improve it?

Terrible plot:
```{r}
flights %>%
  group_by(month, dest) %>%
  summarise(dep_delay = mean(dep_delay, na.rm = TRUE)) %>%
  ggplot(aes(x = factor(month), y = dest, fill = dep_delay)) +
  geom_tile() +
  labs(x = "Month", y = "Destination", fill = "Departure Delay")
```

Futher filtering and grouping to mangle data:
1. Remove NAs
2. Remove departure delays more than 15 mins.
```{r}
flights %>%
  
  group_by(month, dest) %>%
  summarise(dep_delay = mean(dep_delay, na.rm = TRUE)) %>%
  # group by destintation aagin to get number of dests each month
  group_by(dest) %>%
  # only get destinations that have flights in each month of the year
  filter(n() == 12) %>%
  #summarize(n = n())
  # Put back to previous grouped dataframe
  ungroup() %>% 
  # dplyr recorder destination by the 2 values in dest (itself) and then by avg dep_delay
  mutate(dest = reorder(dest, dep_delay)) %>%
  # plot month with factor to categorize value, fill color is affected by dep_delay
  ggplot(aes(x = factor(month), y = dest, fill = dep_delay)) +
  geom_tile() +
  labs(x = "Month", y = "Destination", fill = "Departure Delay")
```




#### Exercise 7.5.3.2

Visualize the distribution of `carat`, partitioned by `price`.

We need to "convert" `price` to a discrete values (binned) so that it can be compared against the discrete values of `carat`
```{r}
# split price into 10 bins and flip for better readability
ggplot(diamonds, aes(x = cut_number(price, 10), y = carat)) +
  geom_boxplot() +
  coord_flip() +
  xlab("Price")
```

```{r}
# split price into increments of $2000
ggplot(diamonds, aes(x = cut_width(price, 2000, boundary = 0), y = carat)) +
  geom_boxplot(varwidth = TRUE) +
  coord_flip() +
  xlab("Price")
```



#### Exercise 7.5.3.3 

How does the price distribution of very large diamonds compare to small diamonds.
Is it as you expect, or does it surprise you?

I don't know much about diamonds but I can see in the chart that the price varies more as the diamonds get bigger, espcially over 2 carats.


