---
output: html_document
editor_options:
  chunk_output_type: console
---
# Exploratory Data Analysis {#exploratory-data-analysis .r4ds-section}

## Introduction {#introduction-3 .r4ds-section}

```{r message=FALSE}
library("tidyverse")
library("viridis")

library("nycflights13")

# This is a cool library
library("lvplot")
```


### Exercise 7.3.1 


Explore the distribution of each of the `x`, `y`, and `z` variables in `diamonds`. What do you learn?
Think about a diamond and how you might decide which dimension is the length, width, and depth.

```{r}

```



### Exercise 7.3.2 


Explore the distribution of price. Do you discover anything unusual or surprising? (Hint: Carefully think about the `binwidth` and make sure you try a wide range of values.)

```{r}

```




### Exercise 7.3.3 

How many diamonds are 0.99 carat?
How many are 1 carat?
What do you think is the cause of the difference?

```{r}

```





#### Exercise 7.5.1.1 

Use what you've learned to improve the visualization of the departure times of cancelled vs. non-cancelled flights.

```{r}
flights = nycflights13::flights 


```


#### Exercise 7.5.1.2


What are the general relationships of each variable with the price of the diamonds?

```{r}

```



#### Exercise 7.5.1.4 

One problem with box plots is that they were developed in an era of much smaller datasets and tend to display a prohibitively large number of "outlying values".
One approach to remedy this problem is the letter value plot.
Install the lvplot package, and try using `geom_lv()` to display the distribution of price vs cut.

An example of the plot is shown below (`geom_lv`). Write answers (not code) to answer these questions:
What do you learn?

How do you interpret the plots?


```{r}
ggplot(diamonds, aes(x = cut, y = price)) +
  geom_lv()
```



#### Exercise 7.5.2.2 

Use `geom_tile()` together with dplyr to explore how average flight delays vary by destination and month of year.
What makes the plot difficult to read?
How could you improve it?

```{r}

```




#### Exercise 7.5.3.2

Visualize the distribution of `carat`, partitioned by `price`.

```{r}

```



#### Exercise 7.5.3.3 

How does the price distribution of very large diamonds compare to small diamonds.
Is it as you expect, or does it surprise you?

```{r}

```

